Individualized treatment effect was predicted best by modeling baseline risk in interaction with treatment assignment

### 2022-03-17 submission/manuscript.tex
update after co-author comments, closes #8

  \begin{abstract}
  \textbf{Objective}: To compare different risk-based methods for
  optimal prediction of individualized treatment effects from RCTs.
  \textbf{Study Design and Setting}: We simulated RCT data using diverse
  assumptions for the average treatment effect, a baseline prognostic
  index of risk (PI), the shape of its interaction with treatment (none,
  linear, quadratic or non-monotonic) and the magnitude of
  treatment-related harms (none or constant independent of the PI). In
  each sample we predicted absolute benefit using: models with a
  constant relative treatment effect; stratification in quarters of the
  PI; models including a linear interaction of treatment with the PI;
  models including an interaction of treatment with a restricted cubic
  spline (RCS) transformation of the PI; an adaptive approach using
  Akaike's Information Criterion. We evaluated predictive performance
  using root mean squared error and measures of discrimination and
  calibration for benefit. \textbf{Results}: The linear-interaction
  model and the RCS-interaction model outperformed the constant
  treatment effect model in many simulation scenarios. The RCS-model was
  optimal when quadratic or non-monotonic deviations from a constant
  treatment effect were stronger, and when sample size was larger.
  Larger sample size also supported the adaptive approach. Illustrations
  in the GUSTO-I trial confirmed these findings. \textbf{Conclusion}: An
  interaction between baseline risk and treatment assignment should be
  considered to improve treatment effect predictions. Non-linear
  interactions should be considered only in larger sample sizes.
  \end{abstract}
   \begin{keyword} treatment effect heterogeneity absolute
benefit prediction models\end{keyword}
 \end{frontmatter}

@@

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions
of benefit, while risk-based methods proved quite robust {[}5{]}. Most
often, risk-modeling approaches are carried out in two steps: first a
risk prediction model is developed externally or internally on the
entire RCT population, ``blinded'' to treatment; then the RCT population
is stratified using this prediction model to evaluate risk-based
treatment effect variation {[}6{]}. This two-step approach identified
substantial absolute treatment effect differences between low-risk and
high-risk patients in a re-analysis of 32 large trials {[}7{]}. However,
even though estimates at the risk subgroup level may be accurate, these
estimates may need further refinement for individual patients,
especially for patients with predicted risk at the boundaries of the
risk intervals. Hence, the risk-stratified approach is useful for
exploring and presenting HTE, but is not sufficient for supporting
treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested
various risk-based models including a prognostic index of baseline risk
(PI) and treatment assignment {[}3,4{]}. We aimed to summarize and
compare different risk-based models for predicting individualized
treatment effects. We simulated RCT settings to compare the performance
of these models under different assumptions of the relation between
baseline risk and treatment. We illustrated the different models by a
case study of predicting individualized effects of treatments for acute
myocardial infarction (MI) in a large randomized controlled trial (RCT).

@@

All risk-based methods assume that a risk prediction model is available
to assign risk predictions to individual patients. For the simulations
we developed a prediction model internally on the entire population,
using a logistic regression model with main effects for all baseline
covariates and treatment assignment. Risk predictions for individual
patients were derived by setting treatment assignment to 0. Another
common approach is to derive the prediction model solely on the control
patients, however this approach has been shown to lead to biased benefit
predictions {[}5,8,9{]}.

@@

For each scenario setting we performed 500 replications, within which
all the considered models were fitted. For comparing between models we
simulated a super-population of size 500,000 for each scenario. We
calculated RMSE and discrimination and calibration for benefit of the
models derived in each replication of our simulation settings within
this super-population. \#\# Empirical illustration

@@

@@

The linear interaction model and the RCS-3 model both displayed very
good performance under many of the considered simulation scenarios, in
contrast with the constant relative treatment effect model. The linear
interaction model was optimal in cases with smaller sample sizes and
moderately performing baseline risk prediction models, that is, it had
lower RMSE, was better calibrated for benefit and had better
discrimination for benefit, even in scenarios with strong quadratic
deviations. In scenarios with true non-monotonic deviations, the linear
interaction model was outperformed by RCS-3, especially in the presence
of true treatment-related harms. Increasing the sample size or the
prediction model's discriminative ability favored RCS-3, especially in
scenarios with non-monotonic deviations and in the presence of
treatment-related harms.

@@

The adaptive approach had adequate performance, following closely on
average the performance of the ``true'' model in most scenarios. With
smaller sample sizes it tended to miss the treatment-risk interactions
and selected simpler models (Supplement Section 4). This conservative
behavior resulted in increased RMSE variability in these scenarios,
especially in the case of true strong linear or non-monotonic deviations
from the base case scenario. Therefore, in the case of smaller sample
sizes the simpler linear interaction model may be a safer choice for
predicting absolute benefits, if a non-constant treatment effect is
suspected.

Risk-based approaches to predictive HTE estimate treatment benefit as a
function of baseline risk. A limitation of our study is that we assumed
treatment benefit to be a function of baseline risk in the majority of
the simulation scenarios. We also considered constant moderate and
strong treatment-related harms, applied on the absolute scale to expand
the range of scenarios in line with previous work {[}15{]}. In a limited
set of scenarios where we assumed the existence of true
treatment-covariate interactions, our conclusions remained unchanged.
Even though the average error rates increased for all the considered
methods, due to the miss-specification of the outcome model, the linear
interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased
results, espcially in the presence of moderate or strong
treatment-related harms. All the results of these simulations can be
found in Supplement, Section 7. Future simulation studies could explore
the effect of more extensive deviations from risk-based treatment
effects.

In our simulations we only focused on risk-based methods, using baseline
risk as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}16--19{]}. Many of these methods rely on
incorporating treatment-covariate interactions in the prediction of
benefit. An important caveat of such approaches is that they may be
prone to overfitting, thus exaggerating the magnitude of the predicted
benefits. In a wide range of simulation settings, a simpler risk
modeling approach was consistently better calibrated for benefit
compared to more complex treatment effect modelling approaches {[}5{]}.
Similarly, when SYNTAX score II, a model developed for identifying
patients with complex coronary artery disease that benefit more from
percutaneous coronary intervention or from coronary artery bypass
grafting was redeveloped using fewer treatment-covariate interactions
had better external performance compared to its predecessor{[}20,21{]}.

>>>>

  \begin{abstract}
  \textbf{Objective}: To compare different risk-based methods for
  optimal prediction of individualized treatment effects. \textbf{Study
  Design and Setting}: We simulated RCT data using diverse assumptions
  for the average treatment effect, a baseline prognostic index of risk
  (PI), the shape of its interaction with treatment (none, linear,
  quadratic or non-monotonic), and the magnitude of treatment-related
  harms (none or constant independent of the PI). We predicted absolute
  benefit using: models with a constant relative treatment effect;
  stratification in quarters of the PI; models including a linear
  interaction of treatment with the PI; models including an interaction
  of treatment with a restricted cubic spline (RCS) transformation of
  the PI; an adaptive approach using Akaike's Information Criterion. We
  evaluated predictive performance using root mean squared error and
  measures of discrimination and calibration for benefit.
  \textbf{Results}: The linear-interaction model and the RCS-interaction
  displayed robust performance across many simulation scenarios. The
  RCS-model was optimal when quadratic or non-monotonic deviations from
  a constant treatment effect were stronger, and when sample size was
  larger. The adaptive approach required larger sample sizes.
  Illustrations in the GUSTO-I trial confirmed these findings.
  \textbf{Conclusion}: An interaction between baseline risk and
  treatment assignment should be considered to improve treatment effect
  predictions.
  \end{abstract}
   \begin{keyword} treatment effect heterogeneity absolute
benefit prediction models\end{keyword}
 \end{frontmatter}

@@

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions
of absolute benefit, while risk-based methods proved quite robust in
terms of benefit calibration, although provided weaker discrimination of
benefit in the presence of true effect modifiers {[}5{]}. Most often,
risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, ``blinded'' to treatment; then the RCT population is
stratified using this prediction model to evaluate risk-based treatment
effect variation {[}6{]}. This two-step approach identified substantial
absolute treatment effect differences between low-risk and high-risk
patients in a re-analysis of 32 large trials {[}7{]}. However, even
though estimates at the risk subgroup level may be accurate, these
estimates may need further refinement for individual patients,
especially for patients with predicted risk at the boundaries of the
risk intervals. Hence, the risk-stratified approach is useful for
exploring and presenting HTE, but is not sufficient for supporting
treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested
various risk-based models including a prognostic index of baseline risk
(PI) and treatment assignment {[}3,4{]}. In the current simulation
study, we aim to summarize and compare different risk-based models for
predicting individualized treatment effects. We simulate different
relations between baseline risk and treatment effects and also consider
potential harms of treatment. We illustrate the different models by a
case study of predicting individualized effects of treatment for acute
myocardial infarction (MI) in a large randomized controlled trial (RCT).

@@

All risk-based methods assume that a risk prediction model is available
to assign risk predictions to individual patients. For the simulations
we developed a prediction model internally on the entire population,
using a logistic regression model with main effects for all baseline
covariates and treatment assignment. Baseline risk predictions for
individual patients were derived by setting treatment assignment to 0.
Another common approach is to derive the prediction model solely on the
control patients, however this approach has been shown to lead to biased
benefit predictions {[}5,8,9{]}.

@@

For each scenario setting we performed 500 replications, within which
all the considered models were fitted. For comparing between models we
simulated a super-population of size 500,000 for each scenario. We
calculated RMSE and discrimination and calibration for benefit of the
models derived in each replication of our simulation settings within
this super-population.

@@

\hypertarget{empirical-illustration-1}{%
\subsection{Empirical illustration}\label{empirical-illustration-1}}

@@

The linear interaction model and the RCS-3 model both displayed very
good performance under many of the considered simulation scenarios. The
linear interaction model was optimal in cases with smaller sample sizes
and moderately performing baseline risk prediction models, that is, it
had lower RMSE, was better calibrated for benefit and had better
discrimination for benefit, even in scenarios with strong quadratic
deviations. In scenarios with true non-monotonic deviations, the linear
interaction model was outperformed by RCS-3, especially in the presence
of true treatment-related harms. Increasing the sample size or the
prediction model's discriminative ability favored RCS-3, especially in
scenarios with non-monotonic deviations and in the presence of
treatment-related harms.

@@

The adaptive approach had adequate performance, following closely on
average the performance of the ``true'' model in most scenarios. With
smaller sample sizes it tended to miss the treatment-risk interactions
and selected simpler models (Supplement Section 4). This conservative
behavior resulted in increased RMSE variability in these scenarios,
especially in the case of true strong linear or non-monotonic deviations
from the base case scenario. Therefore, in the case of smaller sample
sizes the simpler linear interaction model may be a safer choice for
predicting absolute benefits in the presence of any suspected
treatment-related harms.

A limitation of our study is that we assumed treatment benefit to be a
function of baseline risk in the majority of the simulation scenarios.
We also considered constant moderate and strong treatment-related harms,
applied on the absolute scale to expand the range of scenarios in line
with previous work {[}15{]}. In a limited set of scenarios where we
assumed the existence of true treatment-covariate interactions, our
conclusions remained unchanged. Even though the average error rates
increased for all the considered methods, due to the miss-specification
of the outcome model, the linear interaction model had the lowest error
rates. RCS-3 had very comparable performance. The constant treatment
effect model often gave biased results, espcially in the presence of
moderate or strong treatment-related harms. All the results of these
simulations can be found in Supplement, Section 7. Future simulation
studies could explore the effect of more extensive deviations from
risk-based treatment effects.

In our simulations we only focused on risk-based methods, using baseline
risk as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}16--19{]}. Many of these methods rely on
incorporating treatment-covariate interactions in the prediction of
benefit. An important caveat of such approaches is that they may be
prone to overfitting, thus exaggerating the magnitude of the predicted
benefits. In a wide range of simulation settings, a simpler risk
modeling approach was consistently better calibrated for benefit
compared to more complex treatment effect modelling approaches {[}5{]}.
However, whether this remains the case in a range of empirical settings
still needs to be explored Similarly, when SYNTAX score II, a model
developed for identifying patients with complex coronary artery disease
that benefit more from percutaneous coronary intervention or from
coronary artery bypass grafting was redeveloped using fewer
treatment-covariate interactions had better external performance
compared to its predecessor{[}20,21{]}.
